---
title: "final_R_reference.Rmd"
author: "Jackie Stephens"
date: "`r Sys.Date()`"
output: html_document
---

# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages

**Both the function library() and require() load a package into memory**
library() re-loads a package
require() checks first if it is already loaded (less time intensive)
```{r}
#install packages with the name of the package in quote marks if you do not already have the base package
#install.packages("here") 
#install.packages("palmerpenguins")

require(here)
require(palmerpenguins)

```


**2021 Ginkgo Data:** 
```{r}
ginkgo = read.csv(here("data","ginkgo_data_2021.csv"))
#here() calls out where to get the file using quotation marks as well as the .csv at the end
#read.csv() reads the file into a data.frame
#setting it equal to ginkgo= gives the data frame a name
```


## Data Structure

**c() combines arguments into a vector (1 dimensional data consisting of one or more elements of the same type (numbers or characters)).** 
```{r}
## Create a vector of numbers
num_vec  = c(1, 4, 8, 9, 13,17)

## Create a vector of characters
char_vec = c("a", "fish", "data is cool")
#have to put the character elements in quotation marks

#both typing the name of the vector into the console as well as using the print() prints the contents

num_vec
print(num_vec)

char_vec
print(char_vec)
```

**length()**
```{r}
length(num_vec)#length or count of elements in the vector
length(char_vec)#length or count of elements in the vector
```

**matrix()**
```{r}
#creating a matrix from the vector of numbers
print(num_vec)
num_vec_matrix1 = matrix(num_vec, nrow = 2, ncol = 3, 
                         byrow = TRUE,
                        dimnames = list(c("R1","R2"), c("C1","C2","C3")))
num_vec_matrix2 = matrix(num_vec, nrow = 2, ncol = 3, 
                         byrow = FALSE,
                         dimnames = list(c("R1","R2"),c("C1","C2","C3")))

print(num_vec_matrix1)
print(num_vec_matrix2)
#after printing both, you can see that the byrow= argument controls whether elements will be listed vertically (TRUE) or horizontally (FALSE)

#dimnames = names both your rows and columns
```

**data.frame()**
```{r}
#creates data frames of vectors of same lengths
num_vec1 = c(5,6,10,6,3,9,8,5)
num_vec2 = rep(1:2, 4)#repeat 1 to 2 four times
length(num_vec1) 
length(num_vec2)
data.frame(num_vec1,num_vec2)

```

**Examples using the Ginkgo data frame**
```{r}
nrow(ginkgo)#number of rows or observations
ncol(ginkgo)#number of columns
dim(ginkgo)#dimensions of the data frame: rows columns 
```

## Subsetting
```{r}
# use the $ to select a specific column of the data frame
notch_depth = ginkgo$notch_depth

#use [] to subset by position within the data frame

ginkgo[1,]#calls out the entire row of the first observation

ginkgo[2,3]#first number selects the row, second number selects the column

ginkgo[,3]#calls out the third entire column


#use subset() to create a data frame of just adelie species
dat_adelie = subset(penguins, species == "Adelie")
#subset(data frame, column name == "specific value")
```

## Numerical Data exploration
```{r}
#the summary() function will give you the minimum, median, mean, maximum, and 1st and 3rd quantile values of each column
summary(ginkgo)
#because seeds_present was a character value it counts the total FALSE and TRUEs present
summary(penguins)

#the mean() function only accepts numeric or logical values
mean(ginkgo$max_width, na.rm = TRUE)
mean(penguins$body_mass_g, na.rm = TRUE) #adding the na.rm = argument may be necessary if your column contains NAs
#should missing values be removed?

#the standard deviation function 
sd(ginkgo$max_width, na.rm = TRUE)
sd(penguins$body_mass_g,na.rm = TRUE)
```

## Graphical Data Exploration
**Scatterplot**
```{r}
plot(x = ginkgo$max_depth,y = ginkgo$max_width, 
     main = "Maximum Leaf Depth by Leaf Width In Ginkgos",
     xlab = "Maximum Leaf Depth",
     ylab = "Maximum Leaf Width",
     ylim = c(10, 150),
     xlim = c(10,150),
     pch = 1, 
     cex = 0.75, 
     col = "green")

#pch = 0-26 different symbols or shapes 
#cex = number indicating the amount by which plotting text and symbols should be scaled relative to the default. 1=default, 1.5 is 50% larger, 0.5 is 50% smaller
```

**Histogram**
```{r}
summary(penguins$flipper_length_mm)

hist(penguins$flipper_length_mm,
     main = "Distribution of Flipper Length", 
     xlab = "Flipper Length (mm)",
     breaks =  10)
```

**Conditional Boxplots using the Ginkgo Data**
```{r fig.height= 10, fig.width=10}
par(mfrow = c(2, 2))#the c(2,2) means 2 rows of 2 columns for plots
boxplot(ginkgo$petiole_length,
        main = "Ginkgo Petiole Length",
        ylab = "Petiole Length (mm)")
boxplot(ginkgo$max_width,
        main = "Ginkgo Max Width",
        ylab = "Max Width (mm)")
boxplot(ginkgo$petiole_length ~ ginkgo$seeds_present,
        main = "Petiole Length Conditioned By \nSeed Presence",
        xlab = "Seed Present",
        ylab = "Max Width (mm)")
boxplot(ginkgo$max_width ~ ginkgo$seeds_present,
        main = "Max Width Conditioned By Seed Presence",
        xlab = "Seed Present",
        ylab = "Max Width (mm)")
```

## Distribution Functions

The sum of all events in the sample space is 1.

**Normal distribution for continuous:**
use the term density not mass
```{r}
#If mean or sd are not specified they assume the default values of 0 and 1

#probability density function
#is a value of 4 or 5 more likely?
#dnorm(x, mean = 0, sd = 1, log = FALSE)
dnorm(5, mean = 10, sd = 2.5)
dnorm(4, mean = 10, sd = 2.5)

#cumulative density function
#What is the prob of observing 5 or less?
#pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
pnorm(5, mean = 10, sd = 2.5) #prob of 5 or less
1 - pnorm(5, mean = 10, sd = 2.5) #prob of five or more

#quantile function
#what is the 20th percentile of lengths?
#qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
qnorm(.2, mean = 10, sd = 2.5)
```


**Binomial distribution for discrete: two parameters**
1. n: number of trials = size
2. p: prob of success = prob
use the term mass not density
```{r}
#10 plots and from earlier observations 1/4 chance of success

#probability mass function
#dbinom(x, size, prob, log = FALSE)
#what is the prob of observing a value of exactly 4?
dbinom(4, size =  10, prob = .25)

#cumulative mass function
#what is the prob of observing a value of 4 or less? or more?
#pbinom(q, size, prob, lower.tail = TRUE, log.p = FALSE)
pbinom(4, size =  10, prob = .25)
1-pbinom(4, size =  10, prob = .25)

#quantile function
#what is the median or 50th percentile? 90th percentile?
qbinom(0.5, size = 10, prob = .25)
```


